HealpixNet: Efficient spherical Convolutional Neural Network with Healpix sampling for cosmological applications

### 2018-07-25 scnn.tex
update in section 4

Forunately, both of these computational issues can be overcome by defining the convolution kernel $h$ as a polynomial $h_\theta(\lambda) = \sum_{k=0}^K \theta_k \lambda^k$ of degree $K$ parameterized by $K+1$ coefficients $\theta$. In that specific case, the filtering operation \eqnref{graph_convolution} becomes
\begin{equation} \label{eqn:graph_convolution_polynomial}
	h_\theta(\L) \f =  \U \left(\sum_{k=0}^K \theta_k \bLambda^k \right) \U\trans \f = \sum_{k=0}^K \theta_k \L^k \f,
\end{equation}
an operation requiring $k$ multiplication of the graph Laplacian matrix $\L$. In our case, the Laplacian is extremly sparse ($8-9$ entries per line), making the filtering operation for a fixed $k$ linear in the number of pixel $N_{pix}$.
Moreover, here happens the transition from filtering in the spectral domain to filtering in the vertex domain. The matrix $\L^k$ captures $k$-neighborhoods of each pixel, i.e., $(\L^k)_{ij}$ is non-zero if and only if nodes $v_i$ and $v_j$ are connected by a path of length $k$. 
As such, filtering with a polynomial convolution kernel can be interpreted in the pixel (vertex) domain as a weighted linear combination of the pixels neighboors values. The weights are determined by the filter coefficients $\theta$ and the Laplacian $\L$.
\todo{Better vertex domain explanation. \nati{I gave it a try}}

To make the parallel with classical Euclidean signal-domain convolution, i.e., filtering by multiplying the signal with translated versions of a patch, we here match our filter at the level of neighborhoods\footnote{The 1-neighborhood of a node is the set of nodes which are directly connected to it. The 2-neighborhood is the set of nodes which are connected through paths of length 2.} rather than individual nodes. That is required as there is no notion of ordering in a general graph. For example, in a classical convolution on the grid, the relative ordering of vertices is up, down, left, and right. There is no such ordering on the sphere either, especially if irregularly sampled.

Following \citet{defferrard2016convolutional}, we define our filters as Chebyshev polynomials such as the convolution operation becomes
\begin{equation} \label{eqn:graph_convolution_cheby}
	h_\theta(\tL) \f = \sum_{k=0}^K \theta_k T_k(\tL) \f,
\end{equation}
where $\tL = \frac{2}{\lambda_{max}} \L - \b{I} = -\frac{2}{\lambda_{max}} \D^{-1/2} \W \D^{-1/2}$ is the rescaled Laplacian with eigenvalues in $[-1, 1]$ and $T_k(\cdot)$ is the Chebyshev polynomial of degree $k$, defined by the recursive relation $T_k(\tL) = 2\tL T_{k-1}(\tL) - T_{k-2}(\tL)$, $T_1(\tL) = \tL$, $T_0(\tL) = \b{I}$. While definitions \eqnref{graph_convolution_polynomial} and \eqnref{graph_convolution_cheby} both allow the representation of the same set of filters, Chebyshev polynomials are preferred as they form an orthogonal basis on the interval $[-1, 1]$. Moreover, they factor out $\tL^{k-2}$ from $T_k(\tL)$, e.g., the 3-neighborhood is not mixed with the 5-neighborhood. That results in easier learning when optimizing the parameters $\theta$.
Finally, the coefficients can be computed such as the truncated Chebyshev expansions gives an approximate minimax polynomial of a given filter \citep{hammond2011wavelets}. While that is irrelevant when learning filters, it is useful when smoothing a signal on the sphere by filtering it with an approximated low-pass filter.
\todo{(i) scale of $\theta_k$: they are mostly uniform for Chebyshev polynomials, end exploding for monomials. Though SGD implicitly imposes a uniform prior on the learned parameters.}
\todo{(ii) numerical stability of $\L^k \f$ v.s. $T_k(\L) \f$}
\todo{(iii) in the code, initialize $\theta_k$ with stdev adapted to $F_{in}$ and $K$}

@@

Hierarchical pixelization schemes present a natural coarsening. Indeed, each subdivision divides a cell in an equal number of child sub-cells. Coarsening is the reverse operation: merging the sub-cells toward the goal of summarizing the data supported on them. Merging cells lead to a coarser graph, hence the name coarsening. Coarsening defines $\C(i)$, the set of children of node $v_i$. For the HEALPix subdivision scheme, the number of children is constant, i.e., $| \C(i) | = 4^k \ \forall i$, for some $k$.

Pooling refers to the operation that summarizes the data supported on the merged sub-cells in one parent cell. Given a map $\x \in \R^{N_{pix}}$, pooling defines $\y \in \R^{N'_{pix}}$ such as
\begin{equation} \label{eqn:pooling}
	y_i = f \left( \left\{ x_j : j \in \C(i) \right\} \right), \ \forall i \in [N'_{pix}],
\end{equation}
where $f$ is a function which operates on sets (possibly of varying sizes) and $N_{pix} / {N_{pix}}'$ is the down-sampling factor\footnote{$| \C(i) | = N_{pix} / {N_{pix}}' = (N_{side} / N'_{side})^2 = 4^k$ for HEALPix.}. That operation is often taken to be the maximum value, but it might be any permutation invariant operation such as the sum or the average. \figref{pooling} illustrates the process.

@@

The softmax layer is the last layer in a neural network engineered for classification. Given the output $\x \in \R^{N_{classes}}$ of the last FC (or ST), called the logits in the deep learning literature, the softmax layer outputs $\y = SM(\x)$ such that
\begin{equation*}
	y_i = \frac{\exp(x_i)}{\sum_{j=1}^{N_{classes}} \exp(x_j)} \in [0, 1], \ \forall i \in [N_{classes}],
\end{equation*}
where $N_{classes}$ is the number of classes to discriminate. Thanks to the softmax, the output $\y \in \R^{N_{classes}}$ of a neural network is a probability distribution over the classes, i.e., $y_i$ is the probability that the input sample belongs to class $i$. That last layer is actually normalizing $\x$ into $\y$ such that $\| \y \|_1 = \sum_i y_i = 1$.

\todo{Notation GC32 - P256 - FC10 to describe an architecture? Maybe more useful in Experiments where we describe the architecture used for each experiment.}

@@

>>>>

Forunately, both of these computational issues can be overcome by defining the convolution kernel $h$ as a polynomial $h_\theta(\lambda) = \sum_{k=0}^K \theta_k \lambda^k$ of degree $K$ parameterized by $K+1$ coefficients $\theta$. In that specific case, the filtering operation \eqnref{graph_convolution} becomes
\begin{equation} \label{eqn:graph_convolution_polynomial}
	h_\theta(\L) \f =  \U \left(\sum_{k=0}^K \theta_k \bLambda^k \right) \U\trans \f = \sum_{k=0}^K \theta_k \L^k \f,
\end{equation}
an operation requiring $k$ multiplication of the graph Laplacian matrix $\L$. 
In our case, the Laplacian is extremly sparse ($8-9$ entries per line), making the filtering operation for a fixed $K$ linear in the number of pixel $N_{pix}$.
Moreover, here happens the transition from filtering in the spectral domain to filtering in the vertex domain. The matrix $\L^k$ captures $k$-neighborhoods of each pixel, i.e., $(\L^k)_{ij}$ is non-zero if and only if nodes $v_i$ and $v_j$ are connected by a path of length $k$. 
As such, filtering with a polynomial convolution kernel can be interpreted in the pixel (vertex) domain as a weighted linear combination of the pixels neighbors values. The weights are determined by the filter coefficients $\theta$ and the Laplacian $\L$. 
We note that restricting the graph convolutional kernel to a polynomial is similar to restricting the classical Euclidean signal-domain convolution to a patch. In the classical setting, the convolution with a patch is simply a weighted sum of the neighbors with the weights defined by the patch. Similarly, each lines of the matrix $\sum_{k=0}^K \theta_k \L^k$ defines an "irregular patch of radius $K$" for our irregular domain.
\todo{Better vertex domain explanation. \nati{I gave it a try. Should we take a single vertex and explain what heppens to it? Note that I suppressed the following paragraph. }}

% To make the parallel with classical Euclidean signal-domain convolution, i.e., filtering by multiplying the signal with translated versions of a patch, we here match our filter at the level of neighborhoods\footnote{The 1-neighborhood of a node is the set of nodes which are directly connected to it. The 2-neighborhood is the set of nodes which are connected through paths of length 2.} rather than individual nodes. That is required as there is no notion of ordering in a general graph. For example, in a classical convolution on the grid, the relative ordering of vertices is up, down, left, and right. There is no such ordering on the sphere either, especially if irregularly sampled.

Following \citet{defferrard2016convolutional}, we define our filters as Chebyshev polynomials such as the convolution operation becomes
\begin{equation} \label{eqn:graph_convolution_cheby}
	h_\theta(\tL) \f = \sum_{k=0}^K \theta_k T_k(\tL) \f,
\end{equation}
where $\tL = \frac{2}{\lambda_{max}} \L - \b{I} = -\frac{2}{\lambda_{max}} \D^{-1/2} \W \D^{-1/2}$ is the rescaled Laplacian with eigenvalues in $[-1, 1]$ and $T_k(\cdot)$ is the Chebyshev polynomial of degree $k$, defined by the recursive relation $T_k(\tL) = 2\tL T_{k-1}(\tL) - T_{k-2}(\tL)$, $T_1(\tL) = \tL$, $T_0(\tL) = \b{I}$. While definitions \eqnref{graph_convolution_polynomial} and \eqnref{graph_convolution_cheby} both allow the representation of the same set of filters, Chebyshev polynomials are preferred as they form an orthogonal basis on the interval $[-1, 1]$. Moreover, they factor out $\tL^{k-2}$ from $T_k(\tL)$, e.g., the 3-neighborhood is not mixed with the 5-neighborhood. That results in easier learning when optimizing the parameters $\theta$.
Finally, the coefficients can be computed such as the truncated Chebyshev expansions gives an approximate minimax polynomial of a given filter \citep{hammond2011wavelets}. While that is irrelevant when learning filters, it is useful when smoothing a signal on the sphere by filtering it with an approximated low-pass filter.
In practice, we found out that Chebyshev polynomials are slightly more stable in our experiment. We believe that this is due to their orthogonality (as explained above) and uniformity\footnote{The amplitude of Chebyshev polynomials is more or less constant over the domain $[-1, 1]$ independently of the order $k$. On the contrary the amplitude of $x^k$ is very different for $|x|\approx 0$ and $|x| \approx 1$.}.
\nati{I added a sentence and I think we should not give much more details here. If we do an experiment, we can comment in the experiment section.}
\todo{(i) scale of $\theta_k$: they are mostly uniform for Chebyshev polynomials, end exploding for monomials. Though SGD implicitly imposes a uniform prior on the learned parameters.}
\todo{(ii) numerical stability of $\L^k \f$ v.s. $T_k(\L) \f$}
\todo{(iii) in the code, initialize $\theta_k$ with stdev adapted to $F_{in}$ and $K$}

@@

Hierarchical pixelization schemes present a natural coarsening. Indeed, each subdivision divides a cell in an equal number of child sub-cells. Coarsening is the reverse operation: merging the sub-cells toward the goal of summarizing the data supported on them. Merging cells lead to a coarser graph, hence the name coarsening. Coarsening defines $\C(i)$, the set of children of node $v_i$. For the HEALPix subdivision scheme, the number of children is constant, i.e., $| \C(i) | = 4^p \ \forall i$, for some $p$.

Pooling refers to the operation that summarizes the data supported on the merged sub-cells in one parent cell. Given a map $\x \in \R^{N_{pix}}$, pooling defines $\y \in \R^{N'_{pix}}$ such as
\begin{equation} \label{eqn:pooling}
	y_i = f \left( \left\{ x_j : j \in \C(i) \right\} \right), \ \forall i \in [N'_{pix}],
\end{equation}
where $f$ is a function which operates on sets (possibly of varying sizes) and $N_{pix} / {N_{pix}}'$ is the down-sampling factor\footnote{$| \C(i) | = N_{pix} / {N_{pix}}' = (N_{side} / N'_{side})^2 = 4^p$ for HEALPix.}. That operation is often taken to be the maximum value, but it might be any permutation invariant operation such as the sum or the average. \figref{pooling} illustrates the process.

@@

The softmax layer is the last layer in a neural network engineered for classification. Given the output $\x \in \R^{N_{classes}}$ of the last FC (or ST), called the logits in the deep learning literature, the softmax layer outputs $\y = SM(\x)$ such that
\begin{equation*}
	y_i = \frac{\exp(x_i)}{\sum_{j=1}^{N_{classes}} \exp(x_j)} \in [0, 1], \ \forall i \in [N_{classes}],
\end{equation*}
where $N_{classes}$ is the number of classes to discriminate. Thanks to the softmax, the output $\y \in \R^{N_{classes}}$ of a neural network is a probability distribution over the classes, i.e., $y_i$ is the probability that the input sample belongs to class $i$. This last layer is actually normalizing $\x$ into $\y$ such that $\| \y \|_1 = \sum_i y_i = 1$.

@@

\nati{4.8 is long... Should we go in such details?}
\todo{Notation GC32 - P256 - FC10 to describe an architecture? Maybe more useful in Experiments where we describe the architecture used for each experiment.}
